---
title: "DATS 6101 Midterm"
authors: "Brandon Chin, Paul Kelly, Ksenia Shadrina, Luke Wu"
date: "2/16/22"
# date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "show", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

### Potential SMART questions:

1. Were approved mortgage applications from white women under 35 awarded interest rates of one percentage point or higher than those from white men under 35 for properties worth over $400,000 in the United States in 2019?
2. Predict the interest rates for applications of 
### Potential issues with the data:

The smaller sample we selected (the first 10,000 records) is heavily weighted to the state of California. This could skew data unreasonably if we choose a research question that touches upon "state." Should we figure out how to "randomize" selection of dataset rows, or just make a note of it in our EDA?

### Load in dataset

```{r}
hmda <- data.frame(read.csv("2019_lar.csv"))
hmda
```

# Check number of rows

```{r}
nrow(hmda)
```


# Check head and tail of dataframe

```{r}
xkabledplyhead(hmda,5)
xkabledplytail(hmda,5)
```
# view summary and structure
```{r}
summary(hmda)
str(hmda)
```
# Some summaries

```{r}
xkablesummary(hmda[22], title = "Loan amount summary of selected statistics" )
```


#select variables
```{r}
df.selected <- subset(hmda,select=c(derived_ethnicity,derived_race,derived_sex,action_taken,loan_amount,interest_rate,property_value,applicant_age,income,denial_reason_1))
```
# check selected variables' structure
```{r}
str(df.selected$derived_ethnicity)
str(df.selected$derived_race)
str(df.selected$applicant_age)
```

# check missing value
```{r}
missvalue <- is.na(df.selected)
summary(missvalue)
```
# visualization  

```{r}
loadPkg("tidyverse")
ggplot(hmda,aes(x=applicant_sex))+
  geom_boxplot(color="black", fill="purple")+
  labs(title="Boxplot of Applicant Sex", x="Applicant Sex", y="Frequency")
df <- data.frame(id=LETTERS[1:4], min=c(13,15,23,2), max=c(20,30,40,11))

ageFactor <- as.factor(df.selected$applicant_age)
ethnicityFactor <- as.factor(df.selected$derived_ethnicity)
numericrate <- as.numeric(df.selected$interest_rate)
str(df.selected$interest_rate)
ggplot(df.selected, aes(x = factor(derived_sex))) +
    geom_bar()
ggplot(df.selected, aes(x = factor(applicant_age))) +
    geom_bar()
ggplot(df.selected, aes(x = factor(derived_ethnicity))) +
    geom_bar()
hist(df.selected$action_taken)
hist(df.selected$loan_amount)
hist(df.selected$interest_rate)
hist(df.selected$property_value)
hist(df.selected$income)
hist(df.selected$denial_reason_1)
```

The results of this look pretty weird. What are values 3, 4, and 6? 3 = not provided, 4 = not applicable, 6 = applicant selected both male (1) and female (2). We should remove 3, 4, and 6 or create a subset of only 1 and 2, then turn 1 and 2 into factors?

# Subsetting

```{r}
hmda_m_f <- subset(hmda,applicant_sex<3)
```

# Change applicant_sex vector to factor

```{r}
hmda_m_f[, 'applicant_sex'] <- as.factor(hmda_m_f[, 'applicant_sex'])
summary(hmda_m_f)
```

# Re-do the applicant_sex boxplot

```{r}
ggplot(hmda_m_f,aes(x=applicant_sex))+
  geom_boxplot(color="black", fill="purple")+
  labs(title="Boxplot of Applicant Sex", x="Applicant Sex", y="Frequency")
```