---
title: "DATS 6101 Midterm"
authors: "Brandon Chin, Paul Kelly, Ksenia Shadrina, Luke Wu"
date: "2/16/22"
# date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "show", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

```{r modify_outlierKD2}
#' Original outlierKD function by By Klodian Dhana,
#' https://www.r-bloggers.com/identify-describe-plot-and-remove-the-outliers-from-the-dataset/
#' Modified to have third argument for removing outliers instead of interactive prompt,
#' and after removing outlier, original df will not be changed. The function returns the a df,
#' which can be saved as original df name if desired.
#' Also added QQ-plot in the output, with options to show/hide boxplot, histogram, qqplot.
#' Check outliers, and option to remove them, save as a new dataframe.
#' @param df The dataframe.
#' @param var The variable in the dataframe to be checked for outliers
#' @param rm Boolean. Whether to remove outliers or not.
#' @param boxplt Boolean. Whether to show the boxplot, before and after outliers removed.
#' @param histogram Boolean. Whether to show the histogram, before and after outliers removed.
#' @param qqplt Boolean. Whether to show the qqplot, before and after outliers removed.
#' @return The dataframe with outliers replaced by NA if rm==TRUE, or df if nothing changed
#' @examples
#'   outlierKD2(mydf, height, FALSE, TRUE, TRUE, TRUE)
#'   mydf = outlierKD2(mydf, height, TRUE, TRUE, TRUE, TRUE)
#'   mydfnew = outlierKD2(mydf, height, TRUE)
#' @export
outlierKD2 <- function(df, var, rm=TRUE, boxplt=TRUE, histogram=TRUE, qqplt=TRUE) {
  dt = df # duplicate the dataframe for potential alteration
  var_name <- eval(substitute(var),eval(dt))
  na1 <- sum(is.na(var_name))
  m1 <- mean(var_name, na.rm = T)
  colTotal <- boxplt+histogram+qqplt
  par(mfrow=c(2, max(2,colTotal)), oma=c(0,0,3,0)) # fixed issue with only 0 or 1 chart selected
  if (qqplt) {
    qqnorm(var_name, main = "With outliers")
    qqline(var_name)
  }
  if (histogram) { hist(var_name, main="With outliers", xlab=NA, ylab=NA) }
  if (boxplt) { boxplot(var_name, main="With outliers") }

  outlier <- boxplot.stats(var_name)$out
  mo <- mean(outlier)
  var_name <- ifelse(var_name %in% outlier, NA, var_name)
  if (qqplt) {
    qqnorm(var_name, main = "Without outliers")
    qqline(var_name)
  }
  if (histogram) { hist(var_name, main="Without outliers", xlab=NA, ylab=NA) }
  if (boxplt) { boxplot(var_name, main="Without outliers") }
  
  if(colTotal > 0) {  # if no charts are wanted, skip this section
    title("Outlier Check", outer=TRUE)
    na2 <- sum(is.na(var_name))
    cat("Outliers identified:", na2 - na1, "\n")
    cat("Propotion (%) of outliers:", round((na2 - na1) / sum(!is.na(var_name))*100, 1), "\n")
    cat("Mean of the outliers:", round(mo, 2), "\n")
    m2 <- mean(var_name, na.rm = T)
    cat("Mean without removing outliers:", round(m1, 2), "\n")
    cat("Mean if we remove outliers:", round(m2, 2), "\n")
  }

  # response <- readline(prompt="Do you want to remove outliers and to replace with NA? [yes/no]: ")
  # if(response == "y" | response == "yes"){
  if(rm){
      dt[as.character(substitute(var))] <- invisible(var_name)
      #assign(as.character(as.list(match.call())$dt), dt, envir = .GlobalEnv)
      cat("Outliers successfully removed", "\n")
      return(invisible(dt))
  } else {
      cat("Nothing changed", "\n")
      return(invisible(df))
  }
}

```

### Research Topic:

There is considerable evidence indicating lending disparities throughout the United States. (Steil et. al: 2018). For our research topic, we will explore lending practices in one of the fastest appreciating real estate markets over the past thirty years - the state of California. Specifically, we aim to look at the factors that are associated with denials for non-commercial mortgage loans.

Our SMART question is:

“Which factors drove denials for mortgages in California in 2019?”

To answer this question, we are using the Federal Financial Institutions Examination Council's (FFIEC) Home Mortgage Disclosure Act (HMDA) dataset from 2019, located here: https://ffiec.cfpb.gov/data-publication/dynamic-national-loan-level-dataset/2019. We are  focusing on a subset of data of 10,000 observations from 2019 that we will further filter on California leaving us with 5,196 observations.

Our Github repository address is:https://github.com/brandonchin19/Team3/.

### Load in dataset

```{r}
hmda <- data.frame(read.csv("2019_lar.csv"))
str(hmda)
```

# Check number of rows

```{r}
nrow(hmda)
```


# Check head and tail of dataframe

```{r}
xkabledplyhead(hmda,5)
xkabledplytail(hmda,5)
```

# Subsetting to California Only

```{r}
hmda_ca <- subset(hmda,state_code=="CA")
str(hmda_ca)
```

# Subsetting to California Only; non-business properties; principal residences only

```{r}
hmda_ca <- subset(hmda_ca,business_or_commercial_purpose=="2")
str(hmda_ca)
```


```{r}
hmda_ca <- subset(hmda_ca,occupancy_type=="1")
str(hmda_ca)
```

# Subsetting to only relevant columns

```{r}
hmda_ca_final <- hmda_ca[c(10,11,12,13,22,24,39,46,78)]
str(hmda_ca_final)
```


# Changing vector types

```{r}
hmda_ca_final_1 = hmda_ca_final
hmda_ca_final_1$derived_ethnicity = factor(hmda_ca_final$derived_ethnicity)
hmda_ca_final_1$derived_race = factor(hmda_ca_final$derived_race)
hmda_ca_final_1$derived_sex = factor(hmda_ca_final$derived_sex)
hmda_ca_final_1$action_taken = factor(hmda_ca_final$action_taken)
hmda_ca_final_1$loan_amount = as.numeric(hmda_ca_final$loan_amount)
hmda_ca_final_1$interest_rate = as.numeric(hmda_ca_final$interest_rate)
hmda_ca_final_1$property_value = as.numeric(hmda_ca_final$property_value)
hmda_ca_final_1$income = as.numeric(hmda_ca_final$income)
hmda_ca_final_1$applicant_age = factor(hmda_ca_final$applicant_age)
str(hmda_ca_final_1)
```
# checking missing value
```{r}
missvalue <- is.na(hmda_ca_final_1)
summary(missvalue)
```

# remove/replace missing value
```{r}
loadPkg("tidyverse")
hmda_ca_final_2 <- hmda_ca_final_1$interest_rate[is.na(hmda_ca_final_1$interest_rate)] <- mean(hmda_ca_final_1$interest_rate, na.rm = TRUE)

hmda_ca_final_2 <- hmda_ca_final_1 %>% drop_na(property_value)
hmda_ca_final_2 <- na.omit(hmda_ca_final_1, cols="income")
missvalue1 <- is.na(hmda_ca_final_2)
summary(missvalue1)
```
#descripitve analysis
```{r}

Numerical_var <- subset(hmda_ca_final_2,select=c(loan_amount, income, property_value))
var(Numerical_var)
var(hmda_ca_final_2$loan_amount)
var(hmda_ca_final_2$income)
var(hmda_ca_final_2$property_value)
```
# Histograms and Box Plots
```{r}
library(ggplot2)
ggplot(hmda_ca_final_1,aes(x=loan_amount))+
  geom_histogram(color="black", fill="purple")+
  labs(title="Histogram of Loan Amount", x="Loan Amount in dollars", y="Frequency")

ggplot(hmda_ca_final_1, aes(x = factor(derived_sex))) +
    geom_bar()
ggplot(hmda_ca_final_1, aes(x = factor(applicant_age))) +
    geom_bar()
ggplot(hmda_ca_final_1, aes(x = factor(derived_ethnicity))) +
    geom_bar()
ggplot(hmda_ca_final_1, aes(x = factor(action_taken))) +
    geom_bar()

hist(hmda_ca_final_1$loan_amount)
hist(hmda_ca_final_1$interest_rate)
hist(hmda_ca_final_1$property_value)
hist(hmda_ca_final_1$income)


```

Loan amount distribution does not appear to be normal (mean is greater than the median).

```{r}
library(ggplot2)
ggplot(hmda_ca_final_1,aes(x=interest_rate))+
  geom_histogram(color="black", fill="pink")+
  labs(title="Histogram of Interest Rates", x="Interest Rates", y="Frequency")
```

Interest rate distribution does not appear to be normal (mean is greater than the median).


```{r}
ggplot(hmda_ca,aes(y=income))+
  geom_boxplot(color="black", fill="green")+
  labs(title="Boxplot of Incomes", y="Income in dollars", x="Frequency")
```


```{r}
ggplot(hmda_ca_final_1,aes(y=property_value))+
  geom_boxplot(color="black", fill="blue")+
  labs(title="Boxplot of Property Values", y="Property Values in dollars", x="Frequency")
```

```{r}
library(ggplot2)
qplot(data = hmda_ca_final_1,
      x = loan_amount,
      y = derived_race,
      color = action_taken,
      main="Loan amount and Race by Action taken")
```

Loan amounts and loan originations are higher for white borrowers; loan amounts for other races are lower, and loan rejection rates are higher particularly for Black or African American Borrowers. The lower the loan amount, the higher the likelihood of the loan being rejected?

# Exploring columns that appear to have lots of NAs

```{r}
summary(hmda_ca_final_1$interest_rate)
```


```{r}
summary(hmda_ca_final_1$property_value)
```

```{r}
summary(hmda_ca_final_1$income)
```

```{r}
summary(hmda_ca_final_1$loan_amount)
```


```{r}
qqnorm(hmda_ca_final_1$interest_rate,
       main="QQ Plot of Interest Rates",
       ylab="Interest Rate",
       col="pink")
qqline(hmda_ca_final_1$interest_rate)
```


```{r}
qqnorm(hmda_ca_final_1$property_value,
       main="QQ Plot of Property Values",
       ylab="Property Value",
       col="blue")
qqline(hmda_ca_final_1$property_value)
```

```{r}
qqnorm(hmda_ca_final_1$income,
       main="QQ Plot of Income",
       ylab="Income",
       col="green")
qqline(hmda_ca_final_1$income)
```


```{r}
qqnorm(hmda_ca_final_1$loan_amount,
       main="QQ Plot of Loan Amount",
       ylab="Loan Amount",
       col="purple")
qqline(hmda_ca_final_1$loan_amount)
```

# Removing outliers from interest_rate, property value, income, and loan amount

Can we approach more normalized distributions in loan amount, interest rates, income, and property value by removing outliers?

```{r}
hmda_ca_final_no_outliers_1 <- outlierKD2(hmda_ca_final_1,interest_rate)
```


```{r}
hmda_ca_final_no_outliers_2 <- outlierKD2(hmda_ca_final_no_outliers_1,property_value)
```

```{r}
hmda_ca_final_no_outliers_3 <- outlierKD2(hmda_ca_final_no_outliers_2,income)
```

```{r}
loans <- outlierKD2(hmda_ca_final_no_outliers_3,loan_amount)
```

```{r}
qqnorm(loans$interest_rate,
       main="QQ Plot of Interest Rates",
       ylab="Interest Rate",
       col="pink")
qqline(loans$interest_rate)
```
```{r}
qqnorm(loans$property_value,
       main="QQ Plot of Property Values",
       ylab="Property Value",
       col="blue")
qqline(loans$property_value)
```


```{r}
qqnorm(loans$income,
       main="QQ Plot of Income",
       ylab="Income",
       col="green")
qqline(loans$income)
```


```{r}
qqnorm(loans$loan_amount,
       main="QQ Plot of Loan Amount",
       ylab="Loan Amount",
       col="purple")
qqline(loans$loan_amount)
```

With outliers removed, the numerical values are more normally distributed and better suited for analysis.